Moral Foundations of U.S. Political News Organizations

### 2019-04-23 pilot_manuscript.Rmd
Form + Updates from emb2 PDF

abstract: |
  The media ecosystem has grown, and political opinions have diverged such that there are competing conceptions of objective truth. Commentators often point to political biases in news coverage as a catalyst for this political divide. The Moral Foundations Dictionary (MFD) facilitates identification of ideological leanings in text through frequency of the occurrence of certain words. Through web scraping, the researchers extracted articles from popular news sources' websites, calculated MFD word frequencies, and identified words' respective valences. This process attempts to uncover news outlets' positive or negative endorsements of certain moral dimensions concomitant with a particular ideology. In Experiment 1, the researchers gathered poltical articles from four sources. They were unable to reveal significant differences in moral or political endorsements, but they solidified the method to be employed in further research. In Experiment 2, the researchers expanded their number of sources to 10 and analyzed articles that pertain to two specific topics: the 2018 confirmation hearings of U.S. Supreme Court Justice Brett Kavanaugh and the partial U.S. Goverment Shutdown of 2018-2019. Once again, no significant differences in moral or political endorsements were found.

@@

These criticisms of Moral Foundations Theory are valid and should be taken into consideration when conducting research with instruments derived from MFT. The current authors argue these criticisms are especially valid when considered alongside questions regarding the Moral Foundations Dictionary they state herein. However, there exists compelling evidence regarding the validity of Moral Foundations Theory, albeit regarding its application solely through the Moral Foundations Questionnaire. This evidence is discussed within a brief explanation of the Moral Foundations Questionnaire and its relationship to the Moral Foundations Dictionary.

@@

In order to capture language's role in moral and political reasoning, @Graham2009 formulated the Moral Foundations Dictionary (MFD) in order to capture moral reasoning and justification as used in speech and text. The MFD is composed of 259 words, with around 50 words assigned to each of the five foundations. The researchers created a preliminary list of words that they believed would be associated with the five foundations. Then, using the Linguistic Inquiry and Word Count [LIWC; @Pennebaker2007] computer program, they analyzed transcripts of liberal and conservative Christian sermons in order to obtain frequencies of the occurrence of words from the researchers' initial list. The researchers manually checked the results from LIWC in order to make sure the results make sense given the contexts and rhetorical devices used in the sermons, as word frequency analysis ignores sentence context. The researchers offered the following example from a Unitarian sermon as a demonstration of ambiguous statements requiring human verification: "Don't let some self-interested ecclesiastical or government authority tell you what to believe, but read the Bible with your own eyes and open your heart directly to Jesus" [@Graham2009]. This sentence added to the *authority/respect* total in LIWC's analysis, but it appears to suggest that one should reject authority in this context. The researchers eliminated this sentence from the *authority/respect* raw count on account of this discrepancy between the use of authority-related words and the speaker's clear intentions [@Graham2009].

@@

Using the *rvest* library in the statistical package *R*, we pulled body text for individual articles from each of the aforementioned sources (identified using CSS language) and compiled them into a dataset [@Wickham2016]. Using this dataset, we identified word count and average word count per source. This process was run once daily starting in February 2018 until March 2018. Starting in mid-March 2018, the process was run twice daily - once in the morning and again in the evening. Data collection was terminated once 250,000 words per source was collected in April 2018.

@@

As seen in Table \@ref(tab:exp1-source-descriptives), the sources are similar in some aspects yet different in others. Valence appears to be slightly positive across all sources. The large standard deviations seem to indicate little to no presence of a difference in valence. 
*The New York Times* published the greatest number of articles as well as total words. *Breitbart* featured the lowest number of articles, and *NPR* the lowest number of total words from all articles. Per individual article, however, *Breitbart* appears to feature the highest average number of words as well as unique words. Once again the standard deviations call into question any apparent differences between sources. Finally, *Fox News* appears to be the most readable source on average, while *The New York Times* is the least readable. This result might be attributable to the greater number of tokens in the average *New York Times* article compared to *Fox News*. As before, the standard deviations for readability hamper one's ability to draw solid conclusions regarding differences in the sources' use of language.

@@

Table \@ref(tab:exp2-source-descriptives-kav) displays the descriptive statistics for sources' writing on the Kavanaugh confirmation hearing. The sources were similar in most basic linguistic aspects, except for number of articles. For example, *Sean hannity* appears to have published only 27 articles while *Breitbart* published 757 articles on this topic. Valence was found to be slightly positive across all sources. *Fox News* produced the most total words with the most tokens on average. *Politico* featured the highest number of types on average. *Rush Limbaugh* was the most readable source on average by grade level while *Slate* was the least readable. The large standard deviations for these statistics, however, preclude conclusions regarding differences in the sources' use of language.

Table \@ref(tab:exp2-source-descriptives-gs) displays descriptive statistics for articles about the partial government shutdown. Like the Kavanaugh hearing, the sources were similar in average valence (slightly positive). Once again, there was variation in the number of articles published by each source on this topic. *Sean Hannity*, *Rush Limbaugh*, and *The Blaze* published fewer than 100 articles while *Fox News* published 1,013 articles. *Fox News* again featured the most total words and mean tokens. *Politico* had the most types on average. For this topic, *Fox News* was the most readable while *Slate* was the least readable. For each statistic, the excessively high standard deviations render any assertions regarding linguistic differences inconclusive on a descriptive level.

@@

Thus, the researchers' attention turned to the instrument used to detect differences. In the present study, the researchers utilized the Moral Foundations Dictionary, a metric developed from Moral Foundations Theory [@Graham2009]. Based upon the results obtained, it might be necessary to investigate alternative instruments that might better elucidate the differences of interest. Likewise, other theoretical perspectives may be better equipped to explain political differences in discourse.

@@

Other instruments could theoretically be developed through a LIWC analysis, including those that would be useful for analyzing political discourse. Researchers could obtain percent occurrence scores for words appearing in certain LIWC categories, which could then be weighted with valence scores from @Warriner2013. The researchers will consider other applications of the LIWC in future studies aiming to analyze political content. James Pennebaker [-@Pennebaker2007], one of the creators of LIWC, has suggested the program could be applied to analyses of cognitive, social, and emotional states. Furthermore, there is some evidence LIWC can be used to detect deception, but the reliability of the program's accuracy in this regard (67%) calls this assertion into question [@Pennebaker2011]. Despite the promising extent of LIWC's abilities, without a firm theoretical foundation for such analyses, any such applications of LIWC would face serious questions as to their relevance, validity, and generalizability. Nevertheless, LIWC analyses may represent a path to further research into political differences in news media as well as theory formation.

@@

While Relational Frame Theory appears to represent a viable option for further research into political news analysis, there are some drawbacks. First, RFT by itself does not purport to elucidate political differences in discourse, and its applications to political research are not yet well established. In other words, RFT represents one method by which to analyze political text, but with the caveat that the categories are not known to represent patterns of political ideology. One of the main strengths of Moral Foundations Theory in the eyes of the researchers was that it was a theory designed to investigate not only moral differences but also political differences. While the results obtained herein may weaken assumptions regarding the MFD's capabilities with political text, Moral Foundations Theory has been repeatedly established as a strong framework for detecting political differences along moral lines.

@@

@@

>>>>

abstract: |
  The media ecosystem has grown, and political opinions have diverged such that there are competing conceptions of objective truth. Commentators often point to political biases in news coverage as a catalyst for this political divide. The Moral Foundations Dictionary (MFD) facilitates identification of ideological leanings in text through frequency of the occurrence of certain words. Through web scraping, the researchers extracted articles from popular news sources' websites, calculated MFD word frequencies, and identified words' respective valences. This process attempts to uncover news outlets' positive or negative endorsements of certain moral dimensions concomitant with a particular ideology. In Experiment 1, the researchers gathered poltical articles from four sources. We were unable to reveal significant differences in moral or political endorsements, but we solidified the method to be employed in further research. In Experiment 2, the researchers expanded their number of sources to 10 and analyzed articles that pertain to two specific topics: the 2018 confirmation hearings of U.S. Supreme Court Justice Brett Kavanaugh and the partial U.S. Goverment Shutdown of 2018-2019. Once again, no significant differences in moral or political endorsements were found.

@@

These criticisms of Moral Foundations Theory are valid and should be taken into consideration when conducting research with instruments derived from MFT. Such criticisms are especially valid when considered alongside questions about the Moral Foundations Dictionary in particular. Nevertheless, there exists compelling evidence supporting the validity of Moral Foundations Theory, albeit regarding its application solely through the Moral Foundations Questionnaire. This evidence is discussed within a brief explanation of the Moral Foundations Questionnaire and its relationship to the Moral Foundations Dictionary in the next section.

@@

In order to capture language's role in moral and political reasoning, @Graham2009 formulated the Moral Foundations Dictionary (MFD) in order to capture moral reasoning and justification as used in speech and text. The MFD is composed of 259 words, with around 50 words assigned to each of the five foundations. @Graham2009 created a preliminary list of words that they believed would be associated with the five foundations. Then, using the Linguistic Inquiry and Word Count [LIWC; @Pennebaker2007] computer program, they analyzed transcripts of liberal and conservative Christian sermons in order to obtain frequencies of the occurrence of words from the researchers' initial list. @Graham2009 manually checked the results from LIWC in order to make sure the results make sense given the contexts and rhetorical devices used in the sermons, as word frequency analysis ignores sentence context. The researchers offered the following example from a Unitarian sermon as a demonstration of ambiguous statements requiring human verification: "Don't let some self-interested ecclesiastical or government authority tell you what to believe, but read the Bible with your own eyes and open your heart directly to Jesus" [@Graham2009]. This sentence added to the *authority/respect* total in LIWC's analysis, but it appears to suggest that one should reject authority in this context. The researchers eliminated this sentence from the *authority/respect* raw count on account of this discrepancy between the use of authority-related words and the speaker's clear intentions [@Graham2009]. The original Moral Foundations Dictionary can be found in the Appendix.

@@

Using the *rvest* library in the statistical package *R*, we pulled body text for individual articles from each of the aforementioned sources (identified using CSS language) and compiled them into a dataset [@Wickham2016]. Using this dataset, we identified word count and average word count per source. This process was completed once daily starting in February 2018 until March 2018. Starting in mid-March 2018, the process was completed twice daily - once in the morning and again in the evening. Data collection was terminated once 250,000 words per source was collected in April 2018.

@@

As seen in Table \@ref(tab:exp1-source-descriptives), the sources are similar in some aspects yet different in others. Valence appears to be slightly positive across all sources. The large standard deviations seem to indicate little to no presence of a difference in valence across sources. 
*The New York Times* published the greatest number of articles as well as total words. *Breitbart* featured the lowest number of articles, and *NPR* the lowest number of total words from all articles. Per individual article, however, *Breitbart* appears to feature the highest average number of words as well as unique words. Once again the standard deviations call into question any apparent differences between sources. Finally, *Fox News* articles had the lowest reading grade level on average, while *The New York Times* ihad the highest. This result might be attributable to the greater number of tokens in the average *New York Times* article compared to *Fox News*. The standard deviations for readability suggest the presence of a diverse array of articles for each source, ranging from low to high reading level. Large standard deviations suggest the sources feature a lot of overlap between them in their representation of scores.

@@

Table \@ref(tab:exp2-source-descriptives-kav) displays the descriptive statistics for sources' writing on the Kavanaugh confirmation hearing. The sources were similar in most basic linguistic aspects, except for number of articles. For example, *Sean Hannity* appears to have published only 27 articles while *Breitbart* published 757 articles on this topic. Valence was found to be slightly positive across all sources. *Fox News* produced the most total words with the most tokens on average. This is likely due to the fact *Fox News* transcribes many of their videos and publishes them in article form. *Politico* featured the highest number of types on average. *Rush Limbaugh* featured the lowest readability score on average by grade level while *Slate* featured the highest grade-level readability score. The large standard deviations for these statistics, however, preclude conclusions regarding differences in the sources' use of language, as there is likely a lot of overlap between sources' use of language.

Table \@ref(tab:exp2-source-descriptives-gs) displays descriptive statistics for articles about the partial government shutdown. Like the Kavanaugh hearing, the sources were similar in average valence (slightly positive). Once again, there was variation in the number of articles published by each source on this topic. *Sean Hannity*, *Rush Limbaugh*, and *The Blaze* published fewer than 100 articles while *Fox News* published 1,013 articles. *Fox News* again featured the most total words and mean tokens, but this is likely due to to the presence of a large amount of video transcriptions that the organization published as articles. *Politico* had the most types on average. For this topic, *Fox News* featured the lowest reading grade level while *Slate* featured the highest reading grade level. For each statistic, the excessively high standard deviations render any assertions regarding linguistic differences inconclusive on a descriptive level due to the aforementioned overlap in sources' language use.

@@

There remains an aspect of language which this method still cannot address: context. Owing to the presence of millions of words in the analyses presented in the current study, the researchers would be unable to check for context and correct endorsement scores accordingly. @Graham2009 actually performed a process of human error-checking for context when formulating the MFD. While this might have helped that team fix mistakes made by the LIWC program (@Pennebaker2007), manual coding of this nature has the potential to introduce an unnacceptable level of bias in the method, as perceptions of context would be dependent upon raters' interpretations and beleif systems.

Thus, the researchers' attention turned to the instrument used to detect differences. In the present study, the researchers utilized the Moral Foundations Dictionary, a metric developed from Moral Foundations Theory [@Graham2009]. Based upon the results obtained, it might be necessary to investigate alternative instruments that could better elucidate the differences of interest. Likewise, other theoretical perspectives may be more equipped to explain political differences in discourse.

@@

Other instruments could theoretically be developed through a LIWC analysis, including those that would be useful for analyzing political discourse. Researchers could obtain percent occurrence scores for words appearing in certain LIWC categories, which could then be weighted with valence scores from @Warriner2013. The researchers will consider other applications of the LIWC in future studies aiming to analyze political content. @Pennebaker2007 has suggested the program could be applied to analyses of cognitive, social, and emotional states. Furthermore, there is some evidence LIWC can be used to detect deception, but the reliability of the program's accuracy in this regard (67%) calls this assertion into question [@Pennebaker2011]. Despite the promising extent of LIWC's abilities, without a firm theoretical foundation for such analyses, any such applications of LIWC would face serious questions as to their relevance, validity, and generalizability. Nevertheless, LIWC analyses may represent a path to further research into political differences in news media as well as theory formation.

@@

While Relational Frame Theory appears to represent a viable option for further research into political news analysis, there are some drawbacks. First, RFT by itself does not purport to elucidate political differences in discourse, and its applications to political research are not yet well established. In other words, RFT represents one method by which to analyze political text by analyzing occurrence of particular frames. However, there exists the caveat that the frame families are not known to represent patterns of political ideology at this time. One of the main strengths of Moral Foundations Theory in the eyes of the researchers was that it is a theory under which endorsement patterns (direction and relationship) could be predicted based upon political affiliation. While the results obtained herein may weaken assumptions regarding the MFD's capabilities with political text, Moral Foundations Theory has been repeatedly established as a strong framework for detecting political differences along moral lines.

@@

CONTEXT!

@@

# Appendix

```{r appendix_mfd, eval = T, include = T}
orig_mfd = read.csv("exp1_data/original_mfd.csv")
####build the table here####
tableprint = matrix(NA, nrow=length(orig_mfd), ncol=5)
colnames(tableprint) = c("Harm/Care", "Fairness/Reciprocity", "Ingroup/Loyalty", "Authority/Respect", "Purity/Sanctity")

#tableprint[ , 1] = unique(final_gs$Source)
tableprint[ , 1] = orig_mfd[,1]
tableprint[ , 2] = orig_mfd[,2]
tableprint[ , 3] = orig_mfd[,3]
tableprint[ , 4] = orig_mfd[,4]
tableprint[ , 5] = orig_mfd[,5]

tableprint = tableprint[c(1,2,3,4,5)]
tableprint = orig_mfd
#change this to match
apa_table(as.data.frame(tableprint),
          note = "Original Moral Foundations Dictionary from Graham et al. (2009)",
          caption = "Moral Foundations Dictionary", 
          col.names = c("Harm/Care", "Fairness/Reciprocity", "Ingroup/Loyalty", "Authority/Respect", "Purity/Sanctity"),
          format = "latex",
          escape = F,
          small = T, 
          align = c("l", rep("c", 6))
          )
```
